Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2023-01-03 response_to_reviewers_R1.Rmd

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Major comments from Reviewer 1

Overall, the paper is now much more comprehensive, complete, and clear; I think it
will serve as a very useful introduction to the field of sports analytics. I have only a few
suggestions and points of confusion to share, which I hope the authors will examine before
submitting the final version. They are all fairly minor.

> Thank you for your feedback.

- If there is still space allowed in the abstract, I think you could consider explicitly
mentioning the three big ideas (expected value of a game state, win probability, and
measures of team strength). This would give readers a better idea of what is to come
and also allow people to find your article more easily if they are interested in one of
those specific topics.

> Thank you for this suggestion. We have added the three ideas explicitly to the abstract.

- In Section 2.1.2, I am a little confused as to what Carroll et. al. did to extend the
previous model of Carter and Machol, since both efforts seem to involve some sort of
linearity assumption.
- Also in Section 2.1.2, it might be worthwhile to briefly define absorbing states, as I for
one didn’t know the definition before looking it up (and as it turns out the concept is
quite simple)
- In Section 2.1.3, you make the comment “they estimate that a better strategy would
be worth about 0.4 wins per season on average, a substantial amount in a season with
only 16 games.” This makes sense, but previously in Section 2.1.2 you mentioned
a potential difference of 3.8 wins per season, yet didn’t comment on the significant
magnitude of that effect. I’m not advocating for a particular change here, but the
authors could consider adding a small comment about the size of the effect Pelechrinis
et. al. estimated.
- At the end of Section 2.1.3, you write “Lopez (2020) also concludes that coaches are
too conservative on 4th down decision making, but the effect may be overstated due
to the way in which to go yardage has been recorded in the past.” It’s unclear to me
whether the overstated effect is the one Lopez estimates in his paper, or if you are
referring to the estimated effects past studies which used less granular data.
- In Section 2.3, you write “NHL coaches do not always employ the optimal strategies”,
but it isn’t clear what those strategies are in reference to the previous discussion. Is
pulling the goalie optimal?
- In the second formula of Section 4.1, the right hand side of the equation has a 2, but
it should be an α.

@@

Overall Summary:

This revised version is a significant improvement over the previous version and the reviewers have addressed my major concerns with the paper with one exception.

Suggested Changes:

- The authors have now made clearer the limitation in scope of this manuscript related to player evaluation. Perhaps, a statement regarding this can be added to the abstract.

> We have a statement to this effect to the abstract.

- I think that parts of the first paragraph of Section 5 regarding biometric data is a bit out of place. This could be moved to Section 7 (Opportunities). Likewise, Section 5 could be retitled something like‘Betting Data and Sports Analytics” or “Betting Data Analysis”.

### 2023-01-03 response_to_reviewers_R1.Rmd

- Figure 6, in my version the images are stacked rather than left/right

> This is an artifact of the attempt to track changes and does not appear in the unannotated manuscript.

- Section 6.2, line 46 This sentence that starts “Consider, for example, “ seems out of place with the rest of the tone of this article

> Thank you for this suggestion. We have reworded this sentence to read more like the rest of the article: "The surfaces plotted in Figure 9 are helpful in contextualizing player tracking data (such as those shown in Figure 3) and would be laborious for each analyst to have to create on their
own."

@@

- Section 7, line 7 or 8, it might be worth mentioning the growth of small conferences like UCSAS that have sprung up, something like “As the field of sports analytics has grown, there has been a proliferation of sports specific and regional sports analytics conferences.”

> Thank you for this suggestion. We have added a similar sentence to this effect: "As the field of sports analytics has grown, a proliferation of sports specific and regional sports analytics
conferences have arisen."

### 2023-01-03 sports_analytics.Rmd
added scope to abstract; #44
reworded sentence; #44

---
title: "Big Ideas in Sports Analytics and Statistical Tools for their Investigation"
authors:
  - name: Benjamin S. Baumer
#    thanks: Use footnote for providing further information about author (webpage, alternative address)---*not* for acknowledging funding agencies. Optional.
    department: Statistical & Data Sciences
    affiliation: Smith College
    location: Northampton, MA 01063
    email: bbaumer@smith.edu
    ORCID: 0000-0002-3279-0516
  - name: Gregory J. Matthews
    department: Mathematics and Statistics
    affiliation: Loyola University Chicago
    location: Chicago, IL 60660
    email: gmatthews1@luc.edu
    ORCID: 0000-0002-8413-5097
  - name: Quang Nguyen
    department: Statistics & Data Science
    affiliation: Carnegie Mellon University
    location: Pittsburgh, PA 15213
    email: nmquang@cmu.edu
abstract: |
  Sports analytics---broadly defined as the pursuit of improvement in athletic performance through the analysis of data---has expanded its footprint both in the professional sports industry and in academia over the past 30 years. 
  In this paper, we connect three big ideas that are common across multiple sports: the expected value of a game state, win probability, and measures of team strength.
  For each, we explore both the shared similarities and individual idiosyncracies of analytical approaches in each sport. 
  While our focus is on the concepts underlying each type of analysis, any implementation necessarily involves statistical methodologies, computational tools, and data sources. 
  Where appropriate, we outline how data, models, tools, and knowledge of the sport combine to generate actionable insights. 
  This paper should serve as a useful overview for anyone becoming interested in the study of sports analytics. 
keywords:
  - sports analytics
  - R packages
  - sports data
  - pairwise comparisons
  - datasets
link-citations: yes
bibliography: [refs.bib, pkgs.bib]
biblio-style: unsrt   
# use apa for WIRE
csl: apa.csl
output: 
  rticles::arxiv_article:
    keep_md: yes
header-includes:
  - \usepackage{amsmath}
---

@@

Player tracking data contains coordinates that reveal player movement, and these coordinates are always understood in context relative to reference points on the field, court, ice, board, or pitch for a particular sport. 
Orienting these points graphically may require drawing a complex set of guidelines that provide that context to readers. 
Thankfully, the \pkg{sportyR} package [@R-sportyR] contains generic playing surfaces for baseball, basketball, curling, American football, ice hockey, soccer, and tennis that can be added to \pkg{ggplot} graphics with a single function call. 
Consider, for example, how much time the surfaces plotted in Figure \ref{fig:sportyR} would save you, and how helpful they would be in contextualizing player tracking data (such as those shown in Figure \ref{fig:epv-demo}).
With the increased availability of player tracking data, this particular tool should see increased usage.

>>>>

---
title: "Big Ideas in Sports Analytics and Statistical Tools for their Investigation"
authors:
  - name: Benjamin S. Baumer
#    thanks: Use footnote for providing further information about author (webpage, alternative address)---*not* for acknowledging funding agencies. Optional.
    department: Statistical & Data Sciences
    affiliation: Smith College
    location: Northampton, MA 01063
    email: bbaumer@smith.edu
    ORCID: 0000-0002-3279-0516
  - name: Gregory J. Matthews
    department: Mathematics and Statistics
    affiliation: Loyola University Chicago
    location: Chicago, IL 60660
    email: gmatthews1@luc.edu
    ORCID: 0000-0002-8413-5097
  - name: Quang Nguyen
    department: Statistics & Data Science
    affiliation: Carnegie Mellon University
    location: Pittsburgh, PA 15213
    email: nmquang@cmu.edu
abstract: |
  Sports analytics---broadly defined as the pursuit of improvement in athletic performance through the analysis of data---has expanded its footprint both in the professional sports industry and in academia over the past 30 years. 
  In this paper, we connect three big ideas that are common across multiple sports: the expected value of a game state, win probability, and measures of team strength.
  For each, we explore both the shared similarities and individual idiosyncracies of analytical approaches in each sport. 
  While our focus is on the concepts underlying each type of analysis, any implementation necessarily involves statistical methodologies, computational tools, and data sources.
  Where appropriate, we outline how data, models, tools, and knowledge of the sport combine to generate actionable insights. 
  We also describe the use of non-performance data and opportunities to share analytical work, but omit an in-depth discussion of individual player evaluation as beyond our scope.
  This paper should serve as a useful overview for anyone becoming interested in the study of sports analytics. 
keywords:
  - sports analytics
  - R packages
  - sports data
  - pairwise comparisons
  - datasets
link-citations: yes
bibliography: [refs.bib, pkgs.bib]
biblio-style: unsrt   
# use apa for WIRE
csl: apa.csl
output: 
  rticles::arxiv_article:
    keep_md: yes
header-includes:
  - \usepackage{amsmath}
---

@@

Player tracking data contains coordinates that reveal player movement, and these coordinates are always understood in context relative to reference points on the field, court, ice, board, or pitch for a particular sport. 
Orienting these points graphically may require drawing a complex set of guidelines that provide that context to readers. 
Thankfully, the \pkg{sportyR} package [@R-sportyR] contains generic playing surfaces for baseball, basketball, curling, American football, ice hockey, soccer, and tennis that can be added to \pkg{ggplot} graphics with a single function call. 
The surfaces plotted in Figure \ref{fig:sportyR} are helpful in contextualizing player tracking data (such as those shown in Figure \ref{fig:epv-demo}) and would be laborious for each analyst to have to create on their own.
With the increased availability of player tracking data, this particular tool should see increased usage.
